# Data Visualisation {#dataviz}

```{r echo = FALSE}
library(PAMLr)
```

## Formatting data

> **Never forget to get rid of periods where ther logger is not on the bird**

```{r}
# make sure the cropping period is in the correct date format
start = as.POSIXct("2016-07-01","%Y-%m-%d", tz="UTC")
end = as.POSIXct("2017-07-01","%Y-%m-%d", tz="UTC")

# Crop the data
PAM_data= cutPAM(hoopoe,start,end)
```


## Quick multiplots

For a quick look at the data, it's possible to use `quickPLOT`. The user can specify which arguments to use, using `measurements`. There's a choice between different combinations of `"pressure"`, `"light"`, `"acceleration"`, `"temperature"` and `"magnetic"`. You can add any parameters from `?plot`, here I illustrate it with `col="cornflowerblue"` and by showing how to  restrict the x-axis limits `xlim` with the date format, to zoom into the post breeding mihration period of a hoopoe

```{r}
par(mar=c(3,4,0.5,0.5))
quickPLOT(hoopoe, col="cornflowerblue",
          measurements = c("pressure", "light", "acceleration"),
          xlim=c(as.POSIXct("2016-08-20","%Y-%m-%d", tz="UTC"),
                 as.POSIXct("2016-09-01","%Y-%m-%d", tz="UTC")))
```

## Interactive timeseries

To have a better overview of the data, it is possible to create interactive `dygraphPAM()` plots which allow the user to compare different measurements recorded by the logger. These might for instance include light, temperature, pressure, activity, pitch and magnetism. 

If you are **working from Rstudio**, this bit of code should be run:

```{r eval = FALSE}
# In Rstudio, it will display in the viewer by default and use a lot of ram, and is better in html
backup_options <- options() 
options(viewer=NULL) # ensure it is viewed in internet browser
dygraphPAM(dta = PAM_data) # plot
options(backup_options) # restore previous viewer settings
```

If you are **working from base R** use this instead:

To save space here we only plot only one variable - pressure .

```{r cache = TRUE}
dygraphPAM(dta = PAM_data, toPLOT = c("pressure")) 
```

The reason there is additional code for Rstudio, is that by default it will open this graphic in the viewer pane and use up a lot of ram. This  additional code allows the user to open this window in a browser instead of r studio, and the file can later be saved as an html file.

With this interactive plot, the user can then zoom in and out of different plots to help get a feel for the data. For instance, this is a great way of seeing changes in the data which might be due to a logger being in a rucksack and no longer on the birds, or to look at how acticity or pressure might look during migration periods.

It is possible to select areas to zoom into by right clicking and highighting certain regions, and to double click to zoom out. All plots are synched to the same time period and have a timeline at the bottom to increase or decrease the time over which the data is observed.

## Sensor images

Actograms are often used to plot activity over time at different hours of the day. However, the same approach can be used to plot any sensor data, not just activity. For simplicity, we name these “sensor images”. Plotting all sensors side by side is an important step for visualising data and developing an understanding of data patterns, and to start thinking about the behaviours that may be driving the observed patterns. __PAMLr__ offers a function `sensorIMG()`for plotting sensor images, which can be implemented as follows.

```{r}
# Create plots with 3 together (mfrow)
par( mfrow= c(1,3), oma=c(0,2,0,6))

par(mar =  c(4,2,4,2))
sensorIMG(PAM_data$acceleration$date, ploty=FALSE,
          PAM_data$acceleration$act, main = "Activity",
          col=c("black",viridis::cividis(90)), cex=1.2, cex.main = 2)

par(mar =  c(4,2,4,2))
sensorIMG(PAM_data$pressure$date, plotx=TRUE, ploty=FALSE, labely=FALSE,
          PAM_data$pressure$obs,  main="Pressure",
          col=c("black",viridis::cividis(90)), cex=1.2, cex.main = 2)

par(mar =  c(4,2,4,2))
sensorIMG(PAM_data$temperature$date, labely=FALSE,
          PAM_data$temperature$obs,  main="Temperature",
          col=c("black",viridis::cividis(90)), cex=1.2, cex.main = 2)


```

## Histograms and three-dimensional scatterplots

Histograms can provide a first impression of whether some of the data may be aggregated and therefore clustered. Indeed, sensor images may not always well-suited for visualising tri-axial data, such as magnetic field or acceleration. By plotting data in three dimensions (hereafter “3D”) using the function `pam3D` it’s possible to find patterns or clusters of datapoints which would not otherwise be apparent in the data. Here we plot magnetic data. 

```r
pam3D(PAM_data$magnetic$gX, PAM_data$magnetic$gY, PAM_data$magnetic$gZ,
       xlab= "X", ylab= "Y", zlab= "Z",
       xlim=c(-3000,3000), ylim=c(-3000,3000), zlim=c(-3000,3000))
```

```{r echo = FALSE, webgl=TRUE}
knitr::knit_hooks$set(webgl = rgl::hook_webgl)
rgl::plot3d(PAM_data$magnetic$gX, PAM_data$magnetic$gY, PAM_data$magnetic$gZ,
       xlab= "X", ylab= "Y", zlab= "Z",
      xlim=c(-3000,3000), ylim=c(-3000,3000), zlim=c(-3000,3000))
# rgl::rglwidget()
```

## Spherical projections

### g-sphere

A _g-sphere_ is a method of visualising tri-axial acceleration data. This involves centering the data and plotting it on a sphere.The function `triACC` allows the user to center this data (as well as calculating  pitch, roll and yaw from the data)

```r
# plot an g-phere
calibration = triACC(dta = PAM_data$magnetic)
pamSPHERE(x = calibration$centered_accx,
          y = calibration$centered_accy,
          z = calibration$centered_accz,
          ptcol = "royalblue",
          ptsize = 0.03,
          linecolor ="orange",
          spherecolor="orange",
          arrows=TRUE)
```
```{r echo = FALSE}
calibration = triACC(dta = PAM_data$magnetic)
```

```{r echo = FALSE}
knitr::knit_hooks$set(webgl = rgl::hook_webgl)

pamSPHERE(x = calibration$centered_accx,
          y = calibration$centered_accy,
          z = calibration$centered_accz,
          ptcol = "royalblue",
          ptsize = 0.03,
          linecolor ="orange",
          spherecolor="orange",
          arrows=TRUE)
rgl::rglwidget()
```


### m-sphere

An _m-sphere_ is a method of visualising tri-axial magnetometer data. This involves centering the data and correcting the data, before  plotting it on a sphere.The function `triMAG` calibrates the data. This provides the animal's bearing.

```r
# plot a m-phere
calibration = triMAG(dta = PAM_data$magnetic)
pamSPHERE(x = calibration$calib_magx,
          y = calibration$calib_magy,
          z = calibration$calib_magz,
          ptcol = "orange",
          ptsize = 0.03,
          linecolor ="royalblue",
          spherecolor="royalblue",
          arrows=TRUE,
          cex=2)
```
```{r echo = FALSE}
calibration = triMAG(dta = PAM_data$magnetic)
```

```{r echo = FALSE}
knitr::knit_hooks$set(webgl = rgl::hook_webgl)
pamSPHERE(x = calibration$calib_magx,
          y = calibration$calib_magy,
          z = calibration$calib_magz,
          ptcol = "orange",
          ptsize = 0.03,
          linecolor ="royalblue",
          spherecolor="royalblue",
          arrows=TRUE,
          cex=2)
rgl::rglwidget()
```
